---
title: "Spotify"
description: |
  Analysis of Spotify from TidyTuesday
author: Phu Vo
date: September 23, 2025
format: html
---

## Introduction
The Spotify dataset contains a file named spotify_songs.csv, and is a collection
of thousands of songs with unique variables such as track_id, track_name, track_artist, and track_popularity, among others. One of the variables I found to be interesting and analyzeable was "danceability." Recently, I've started going to some parties again, but the DJs have been playing pretty bad songs... I decided to find out the artists who produce the "best" party music using the data.
Sources of the TidyTuesday dataset will be provided at the bottom of the page.

## Code and Graph

```{r}
#| include: false
#| message: false
#| warning: false

library(tidyverse)
```

```{r}
#| include: false

# TidyTuesday Spotify Dataset (week 01-21-2020).

spotify <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-01-21/spotify_songs.csv')

```

```{r}
#| echo: true

#Filtered by popular tracks with scores >= 90
popular_tracks <- spotify |>
  filter(track_popularity >= 90)

# Top 10 artists by average danceability among popular tracks
top_danceable_popular <- popular_tracks |>
  group_by(track_artist) |>
  summarize(avg_danceability = mean(danceability, na.rm = TRUE),
            .groups = "drop") |>
  arrange(desc(avg_danceability)) |>
  slice_head(n = 10)

# Plot
ggplot(top_danceable_popular, aes(x = reorder(track_artist, avg_danceability),
                                  y = avg_danceability)) +
  geom_col(fill = "darkseagreen2") +
  geom_text(aes(label = round(avg_danceability, 2)),
            hjust = -.1,
            size = 2) +
  coord_flip() +
  labs(title = "Most Danceable Artists filtered by Popular Tracks >=90 (Frat Party Final Bosses)",
       x = "Artist",
       y = "Average Danceability") +
  theme_light()+
  theme(plot.title = element_text(size = 12),
        axis.text.y = element_text(size = 8)
        )
```

## Analysis
Influenced by party music, I created a horizontal bar chart to show the top 10 artists with tracks that have the highest average danceability. A score of 0 would represent undanceable, and a score of 1 to be extremely danceable. I decided to sort songs by popularities of greater than or equal to 90, because the song has to be "relevant" enough for a college to know and be able to vibe/dance to. Roddy Rich tops the list with the highest average danceability of 0.90 --he's followed by artists like Dimelo Flow, Lil Uzi Vert, Regard, Rauw Alejandro, and Lil Nas X, with their scores at around 0.90. The above artists are Hip/Hop creators, Reggaeton singers, and DJs, which make sense-- the cultural background of the music they create has a strong history of dance and high energy levels. Next, Y2K scored 0.84, Ed Sheeran scored 0.82, Tones and I 0.82, and Nicky Jam 0.82. The results suggest that among mainstream artists, hip-hop, reggaeton, and pop crossovers tend to dominate in term of danceable tracks. It's important to note that filtering an artist by their top hits only may not accurately represent their music's danceability. For example, Ed Sheeran has a lot of indie music on top of his pop hits, however, as they aren't as relevant, the popularity filter does not account for the danceability of those tracks, increasing his score. However, I believe that the parameters still give us a good way to analyze which artists we should be playing to dance. So, the next time I go to a party, or high-octane social event, I better hear one of these artists! Or, we'll just get some 2010s house again...

## Link and Credits
The data comes from Spotify via the spotifyr package, authored by Charlie Thompson, Josiah Parry, Donal Phipps, and Tom Wolff. The link to the dataset can be accessed here: [TidyTuesday Spotify Dataset](https://github.com/rfordatascience/tidytuesday/tree/main/data/2020/2020-01-21).